---
title: "F1_Data"
output: html_document
date: "2024-02-07"
---

Load Package

```{r setup, include=FALSE}
library(f1dataR)
library(dplyr)
library(ggplot2)

round_n <- 12  # Round of the season
season_n <- 2023 # Year of the season
```

Colors for tyres and teams, and relation driver-team

```{r}
library(racinglibraries)

scuderia_colors <- scuderia_colors() # Correlates team and color
tyre_colors <- tyre_colors()  # Correlates tyre and color
drivers_teams <- team_scuderia()  # Correlates driver and team
drivers_code <- codigo_piloto()  # Abbreviation of Driver Names

associate_color <- function(pilotos, colores, equipos) {
  colores_pilotos <- sapply(pilotos, function(piloto) unname(colores[equipos[piloto]]))
  return(colores_pilotos)
}

# Correlates driver and color
drivers_color <- associate_color(names(drivers_teams),
                                 scuderia_colors,drivers_teams) 
```


Study Qualifying performance between teammates

```{r}

# Load Qualy Times and add relevant information

qualifying <- f1dataR::load_quali(season = season_n, round = round_n)

qualifying$driver_id <- drivers_code[match(qualifying$driver_id, names(drivers_code))]
qualifying$Team <- drivers_teams[match(qualifying$driver_id,names(drivers_teams))]


# Use Pole lap as a reference time

pole_lap <- qualifying[1,]$q3_sec

# Calculate Gaps for each Qualy round (Q1,Q2,Q3)

gaps_mates <- qualifying %>%
  group_by(Team) %>%
  summarise(
    fast_driver = first(driver_id),
    dif_q1 = max(q1_sec) - min(q1_sec),
    dif_q2 = max(q2_sec) - min(q2_sec),
    dif_q3 = max(q3_sec) - min(q3_sec)
  )


#If Albon gets into Q2 but Sargeant doesn't, then the gap between them will be
#the difference of their respective fastest times exclusively in Q1.

#If Alonso gets into Q3 but Stroll is eliminated in Q2, then the gap between them 
#will be the difference of their respective fastest times exclusively in Q2.

# Isolate the last round where both teammates were present

gaps_mates  <- gaps_mates %>%
  summarise(Team = Team,
            fast_driver = fast_driver,
            gap = if_else(!is.na(dif_q3), 
                          dif_q3, 
                          if_else(!is.na(dif_q2), dif_q2, dif_q1)),
            perc_gap = round(gap*100/pole_lap,3)) %>%
  arrange(perc_gap)

gaps_mates[,"fast_driver"] <- factor(gaps_mates$fast_driver, 
                                     levels = gaps_mates$fast_driver)
```

```{r}
ggplot(data = gaps_mates, aes(fast_driver,perc_gap, fill = Team)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = scuderia_colors) +
  theme_dark_f1(axis_marks = TRUE) +
  labs(title = "% Gap between Teammates") 
```

--------------------------------------------------------------------------------


Load Results of the race
```{r}

# FInal results of the race
race_result <- f1dataR::load_results(season = season_n,
                                     round = round_n)

race_result$driver <- drivers_code[match(race_result$driver_id,names(drivers_code))]
race_result$driver <- factor(race_result$driver,levels = race_result$driver)
race_result$team <- drivers_teams[match(race_result$driver,names(drivers_teams))]

```


Evaluate Top speeds reached during the race.
```{r}

# Discard NA values of top speeds
speeds <- race_result[!is.na(race_result$top_speed_kph),]
# Sort the speeds
speeds <- speeds %>%
  arrange(desc(top_speed_kph))
# Convert speeds to factor for the plot
speeds$driver <- factor(speeds$driver,levels = speeds$driver)
                            
ggplot(data = speeds, aes(driver,top_speed_kph, fill = team)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = scuderia_colors) +
  theme_dark_f1(axis_marks = TRUE) +
  labs(title = "Top Speeds")
```

Positions Gained and Lost of each driver
```{r}

#Change column names
race_result$lap_number = 0
colnames(race_result)[5] <- "position"
colnames(race_result)[4] <- "final_position"

# Calculate the balance of positions gained
race_result$position <- as.numeric(race_result$position)
race_result$final_position <- as.numeric(race_result$final_position)
race_result[,"balance"] <- race_result$position - race_result$final_position

ggplot(data = race_result[race_result$position != 0,],
       aes(driver,balance,fill = team, color = status)) + 
  geom_bar(stat = "identity", size = 1) +
  scale_fill_manual(values = scuderia_colors) +
  theme_dark_f1(axis_marks = TRUE) +
  labs(title = "Positions Gained/Lost during the Race")
```



Lap by lap Information about the race
```{r}
# Download all the information about the race
race_times <- f1dataR::load_session_laps(season = season_n,
                                round = round_n,
                                session = "R")

race_times$team <- drivers_teams[race_times$driver]
# Arrange stints of drivers
stint <- paste("Stint",race_times$stint,sep = "_")
race_times[,"stint"] <- stint

```



Evolution of track position during the race

```{r}
ggplot(data = race_times,aes(lap_number, position, line = driver, point = driver)) +
  geom_point(data = race_times, aes(lap_number, position, color = team))+
  geom_point(data = race_result, aes(lap_number,position,color = team)) +
  geom_line(data = race_times, aes(lap_number,position,color = team)) +
  geom_line(data = race_result, aes(lap_number,position,color = team)) +
  geom_label(data = race_result[race_result$lap_number == 0,],
             aes(label = driver, fill = team,hjust = 1.2), show.legend = F) +
  geom_label(data = race_times[race_times$lap_number == max(race_times$lap_number),],
             aes(label = driver, fill = team),hjust = -0.15, show.legend = F) +
  scale_color_manual(values = scuderia_colors) +
  scale_fill_manual(values = scuderia_colors) +
  scale_y_reverse() +
  theme_dark_f1(axis_marks = TRUE)
```
Pit stops of each driver

```{r}
pit_stops <- race_times %>%
  group_by(driver,team) %>%
  filter(pit_out_time != "NaN") %>%
  summarise(pit_stops = n())

ggplot(pit_stops,aes(pit_stops,driver,fill = team)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = scuderia_colors) +
  theme_dark_f1(axis_marks = TRUE) +
  labs(title = "Pit Stops Made")
```

BoxPlot of Race Pace
```{r}

# Eliminate drivers which ahve not completed more than 75 % laps

filtered_race_times <- race_times %>%
  mutate(
    total_laps = max(lap_number)
  ) %>%
  group_by(driver,team) %>%
  mutate(
    laps_completed = max(lap_number)
    ) %>%
  filter(laps_completed == total_laps) %>%
  filter(lap_time != "NaN") %>%
  select(-laps_completed,-total_laps)


filtered_race_times$driver <- factor(filtered_race_times$driver,
                                     levels = unique(filtered_race_times$driver))
  
ggplot(data = filtered_race_times, aes(driver,lap_time,fill = team)) +
  geom_boxplot(color = "brown",outlier.color = "white") +
  scale_fill_manual(values = scuderia_colors) +
  theme_dark_f1(axis_marks = TRUE)
```



Pace by stint of each Driver

```{r}

drivers_interest <- filtered_race_times[filtered_race_times$driver_number %in% c(16,14,44,1),]

stints <- drivers_interest %>%
  group_by(driver,stint,compound) %>%
  filter(
    lap_number == min(lap_number)
  ) %>%
  summarise(
    lap_number = lap_number,
    lap_time = lap_time
  )

n_stints <- length(unique(filtered_race_times$stint))

# Full race

ggplot(data = drivers_interest, aes(lap_number, lap_time, color = driver)) +
  geom_point(alpha = 0) +
  geom_line() +
  scale_color_manual(values = drivers_color) +
  scale_fill_manual(values = tyre_colors) +
  geom_label(data = stints, mapping = aes(label = compound, fill = compound), color = "black",show.legend = FALSE) + 
  theme_dark_f1(axis_marks = TRUE) 

# Separate by stints
ggplot(data = drivers_interest, aes(lap_number,lap_time, color = driver, fill = compound)) +
  geom_line() +
  scale_color_manual(values = drivers_color) +
  scale_fill_manual(values = tyre_colors) +
  geom_label(data = stints, mapping = aes(label = compound), color = "black", show.legend = FALSE) + 
  facet_wrap(~stint, ncol = n_stints, scales = "free_x") +
  theme_dark_f1(axis_marks = TRUE) 
```



```{r}

# Eliminate outlier laps

race_wo_outliers <- race_times[race_times$lap_time != "NaN",] %>%
  filter(lap_time < quantile(lap_time, 0.75) + 1.5*IQR(lap_time))


# Who made the average faster lap times
library(tidyverse)
avg_laptimes <- race_wo_outliers %>%
  group_by(driver, team) %>%
  summarise(
    avg_lap = mean(lap_time),
  )

# Fastest driver on average
reference <- min(avg_laptimes$avg_lap)
pace_gaps <- avg_laptimes %>%
  mutate(
    gap_secs = avg_lap - reference
  ) %>%
  arrange(gap_secs)

# Ordenar pilotos

pace_gaps$driver <- factor(pace_gaps$driver, levels = pace_gaps$driver)

ggplot(data = pace_gaps, aes(driver, gap_secs, fill = team)) +
  geom_bar(stat = "identity")  +
  scale_fill_manual(values = scuderia_colors) +
  theme_dark_f1(axis_marks = TRUE) +
  labs(title = "Gap in race pace to fastest")
```


Stints of the driver

```{r}
driver_stints <- "HAM"
ggplot(data = race_wo_outliers[race_wo_outliers$driver == driver_stints,], aes(stint,lap_time,fill = compound) ) +
  geom_boxplot(outlier.colour = "purple", color = "orange") +
  labs(title = paste("Stints",driver_stints,sep = " ")) +
  theme(title = element_text(colour = "white")) +
  theme(axis.text.x = element_blank()) +
  scale_fill_manual(values = tyre_colors) +
  theme_dark_f1(axis_marks = TRUE)
```

Show best strategy

```{r}
strategy <- race_times %>%
  group_by(driver, team,stint,compound) %>%
    summarise(
    time_stint = sum(lap_time),
    lap_race = max(lap_number) - min(lap_number) + 1,
    strategy = paste(unique(compound),collapse = "-")
    ) %>%
  group_by(driver,team) %>%
  summarise(
    strategy = paste(strategy,collapse = "-"),
    compound = compound,
    time_stint = time_stint,
    lap_race = lap_race,
    total_time = sum(time_stint),
    stint = stint
  ) %>%
  arrange(total_time)
```

Msotrar estrategias de los pilotos

```{r}

strategy$compound <- factor(strategy$compound,levels = unique(strategy$compound))
ggplot(race_times, aes(lap_number,y=driver,color=compound,fill = stint)) +
  geom_line(size = 3) +
  scale_color_manual(values = tyre_colors) + 
  theme(axis.text.x = element_text(angle = 90)) +
  theme_dark_f1(axis_marks = TRUE) + 
  scale_x_continuous(breaks = seq(1,max(race_times$lap_number),by=11))
```

Ver la degradación de neumatico

```{r}
deg <- race_wo_outliers[race_wo_outliers$pit_in_time == "NaN",] %>%
  group_by(driver,stint,team,compound) %>%
  summarise(
    dif_time = (median(lap_time) - min(lap_time))/n(),
    n_laps = n()
  )

ggplot(data = deg, aes(n_laps,dif_time,fill = team)) +
  geom_point(alpha = 0) +
  scale_fill_manual(values = scuderia_colors)+
  scale_color_manual(values = tyre_colors) +
  geom_label(aes(label = driver, color = compound), show.legend = T, label.size = 0.9) +
  geom_text(aes(label = driver),color = "purple", show.legend = F) +
  theme_dark_f1(axis_marks = TRUE) +
  facet_wrap(~compound) 
  scale_y_reverse()
```



